---
editor_options:
  markdown:
    wrap: 72
title: Contract Rates - Accessorial Analysis
---

# Load Packages

```{r, echo = f}
library(bigrquery)
library(DBI)
library(tidyverse)
library(stats)
library(ggfortify)
```

# Create BQ connection

```{r}
bqcon <- dbConnect(
  bigrquery::bigquery(),
  project = "freightwaves-data-science"
)
```

# Pull data from BQ

```{r, echo = f}
query1 = "with data as (
    select
      cass_shipment_id
      ,ship_date
      ,shipper_master_code
      ,substring(origin_postal_code, 1, 3) as origin_zip3
      ,substring(destination_postal_code, 1, 3) as dest_zip3
      ,upper(system_type) as system_type
      ,case 
        when transportation_mode_description in ('TRUCKLOAD (DRY VAN)', 'TRUCKLOAD', 'MOTOR', 'BROKER') then 'VAN'
        when transportation_mode_description = 'TRUCKLOAD (TEMP-CONTROLLED)'then 'REEFER'
        else 'OTHER'
        end as mode
      ,case
        when distance < 100 then 'City'
        when distance < 250 then 'Short'
        when distance < 500 then 'Mid'
        when distance < 750 then 'Tweener'
        when distance < 1000 then 'Long'
        when distance < 1500 then 'Extra-long'
        else 'Cross-country'
        end as LOH
      ,case
        when upper(accessorial_charge_description) like '%BASE%' then 'BASE CHARGE'
        when upper(accessorial_charge_description) like '%ENERGY%SUR%(FUEL%' then 'ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)'
        when upper(accessorial_charge_description) like '%ENERGY%CH%(FUEL%' then 'ENERGY CHARGE (FUEL ADJUSTMENT FACTOR)'
        when upper(accessorial_charge_description) like '%ENERGY%SUR%' then 'ENERGY SURCHARGE'
        when upper(accessorial_charge_description) like '%ENERGY%CH%' then 'ENERGY CHARGE'
        when upper(accessorial_charge_description) like '%ENERGY%' then 'ENERGY OTHER'
        when upper(accessorial_charge_description) like '%FUEL%SUR%' then 'FUEL SURCHARGE'
        when upper(accessorial_charge_description) like '%SUR%FUEL%' then 'FUEL SURCHARGE'
        when upper(accessorial_charge_description) like '%FUEL%CH%' then 'FUEL CHARGE'
        when upper(accessorial_charge_description) like '%FUEL%TAX%' then 'FUEL TAX'
        when upper(accessorial_charge_description) like 'FUEL' then 'FUEL'
        when upper(accessorial_charge_description) like '%FUEL%ADJ%' then 'FUEL ADJUSTMENT'
        when upper(accessorial_charge_description) like '%DELIVERY%FUEL%' then 'FUEL DELIVERY'
        when upper(accessorial_charge_description) like '%ORIGIN%FUEL%' then 'FUEL ORIGIN'
        when upper(accessorial_charge_description) like '%DESTINATION%FUEL%' then 'FUEL DESTINATION'
        when upper(accessorial_charge_description) like '%FUEL%' then 'FUEL OTHER'
        else upper(accessorial_charge_description)
        end as accessorial_charge_description
      ,accessorial_charge_amount
      ,amount_paid
    
    from `freightwaves-data-factory.warehouse.beetlejuice`
    
    where (upper(accessorial_charge_description) like '%FUEL%'
        or upper(accessorial_charge_description) like '%ENERGY%'
        or upper(accessorial_charge_description) like '%BASE%')
  )
  
  select
    *
  from data
  where 
    -- mode in ('VAN', 'REEFER') and
    ship_date >= '2021-01-01' and
    ship_date <= current_date"

query2 <- "SELECT * FROM `freightwaves-data-science.dbt_jliebert.accessorial_analysis_query`"

bj <- dbGetQuery(bqcon, query1)
```

# Check for most common accessorials

```{r, echo = F}
accessorial_table <- bj %>% 
  filter(accessorial_charge_description != "BASE CHARGE") %>%
  group_by(accessorial_charge_description, mode) %>% 
  tally() %>%
  arrange(mode, desc(n)) %>%
  print()
```

# Group by month and year
```{r, echo = F}
monyear_table <- bj %>% 
  filter(ship_date <= "2022-09-15") %>%
  filter(accessorial_charge_description %in% c("ENERGY CHARGE", 
                                               "ENERGY SURCHARGE",
                                               "ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)",
                                               "FUEL", "FUEL CHARGE", "FUEL SURCHARGE", "FUEL OTHER")) %>% 
  group_by(year = lubridate::floor_date(ship_date, "year"),
           month = lubridate::floor_date(ship_date, "month"),
           accessorial_charge_description) %>% 
  summarize(count = n()) %>% 
  print()

monyear_plot <- monyear_table %>% 
  ggplot(aes(month, count, fill=accessorial_charge_description)) +
  geom_line(aes(color=accessorial_charge_description)) +
  theme_classic()
monyear_plot
```

# Create base rate column

```{r, echo = F}
base_total <- bj %>%
  group_by(cass_shipment_id) %>%
  filter(!grepl('BASE', accessorial_charge_description)) %>%
  summarize(non_base_total = sum(accessorial_charge_amount, na.rm=T), total = max(amount_paid)) %>%
  mutate(base_rate = total - non_base_total)

all_data <- bj %>%
  left_join(., base_total, by=c('cass_shipment_id')) %>%
  mutate(non_base_total = ifelse(is.na(non_base_total), 0, non_base_total),
         total          = ifelse(is.na(total), amount_paid, total),
         base_rate      = ifelse(is.na(base_rate), amount_paid, base_rate))

good_data <- all_data %>% 
  mutate(lane = paste(origin_zip3, dest_zip3, sep = ':')) %>% 
  filter(base_rate > 0 & base_rate <= 5000 &
           total > 0 & total <= 10000)
```

# Filter out lanes with < 3 carriers

```{r}
volume_lanes <- good_data %>%
  mutate(week = strftime(ship_date, '%V%')) %>%
  group_by(mode, origin_zip3, dest_zip3, LOH, week) %>%
  summarize(Load_Count = n(), Carrier_Count = length(unique(shipper_master_code))) %>%
  filter(Carrier_Count >= 3) %>%
  mutate(lane = paste(origin_zip3, dest_zip3, sep = ':'))

use_data <- good_data %>% filter(lane %in% volume_lanes$lane)
```

# Create summary stats

```{r}
stats_data <- use_data %>% 
  group_by(accessorial_charge_description, mode, origin_zip3, dest_zip3, LOH) %>% 
  summarize(min   = min(accessorial_charge_amount),
            max   = max(accessorial_charge_amount),
            avg   = mean(accessorial_charge_amount),
            med   = median(accessorial_charge_amount),
            var   = var(accessorial_charge_amount),
            range = max(accessorial_charge_amount) - min(accessorial_charge_amount),
            pct_of_base  = median(accessorial_charge_amount/base_rate),
            pct_of_total = median(accessorial_charge_amount/amount_paid),
            count = n()) %>% 
  filter(!grepl('BASE CHARGE', accessorial_charge_description)) %>%
  filter(is.na(origin_zip3) == FALSE & is.na(dest_zip3) == FALSE & var != 0) %>% 
  ungroup() %>% 
  mutate(cd_min = cume_dist(min),
         cd_max = cume_dist(max),
         cd_avg = cume_dist(avg),
         cd_med = cume_dist(med),
         cd_pct_of_base = cume_dist(pct_of_base),
         cd_pct_of_total = cume_dist(pct_of_total),
         ) %>%
  # remove outliers
  filter(cd_min >= 0.1 & cd_min <= 0.9 &
         cd_max >= 0.1 & cd_max <= 0.9 &
         cd_avg >= 0.1 & cd_avg <= 0.9 &
         cd_med >= 0.1 & cd_med <= 0.9 &
         cd_pct_of_base >= 0.1 & cd_pct_of_base <= 0.9 &
         cd_pct_of_total >= 0.1 & cd_pct_of_total <= 0.9) %>% 
#   filter(min >= 0 & min <= 2000 &
#          max >= 0 & max <= 3000 &
#          avg >= 0 & avg <= 2000 &
#          med >= 0 & med <= 2000 &
#          pct_of_base >= 0 & pct_of_base <= 5 &
#          pct_of_total >= 0 & pct_of_total <= 1) %>% 
  select(-c(cd_min, cd_max, cd_avg, cd_med, cd_pct_of_base, cd_pct_of_total)) %>% 
  drop_na()
```
# Scale data

```{r}
# stats_data2 <- stats_data %>% 
#   filter(mode %in% c("VAN", "REEFER"))

scaled_data <- stats_data[,1:5] %>% 
  cbind(scale(stats_data[,6:13]))
```

# Group the stats at different granularities

```{r}
stats_df <- scaled_data %>%
  group_by(accessorial_charge_description) %>%
  summarize(lane_count = n(),
            med_min   = round(median(min), 2),
            med_max   = round(median(max), 2),
            med_avg   = round(median(avg), 2),
            med_med   = round(median(med), 2),
            med_pct_of_base = round(median(pct_of_base), 2),
            med_pct_of_total = round(median(pct_of_total), 2),
            med_range = round(median(range), 2),
            med_var   = round(median(var), 2)
           ) %>% 
  arrange(accessorial_charge_description) %>%
  filter(accessorial_charge_description %in% c("FUEL CHARGE",
                                               # "ENERGY CHARGE", "ENERGY SURCHARGE", "ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)", "FUEL", "FUEL OTHER",
                                               "FUEL SURCHARGE"))

stats_plot <- stats_df[,-2] %>%
  gather(stat, value, med_min:med_var) %>% 
  ggplot(aes(stat, value, fill=accessorial_charge_description)) +
  geom_bar(stat="identity", position="dodge") +
  theme_classic()
stats_plot


stats_df_LOH <- scaled_data %>%
  group_by(LOH, accessorial_charge_description) %>%
  summarize(lane_count = n(),
            med_min   = round(median(min), 2),
            med_max   = round(median(max), 2),
            med_avg   = round(median(avg), 2),
            med_med   = round(median(med), 2),
            med_pct_of_base = round(median(pct_of_base), 2),
            med_pct_of_total = round(median(pct_of_total), 2),
            med_range = round(median(range), 2),
            med_var   = round(median(var), 2)
           ) %>% 
  arrange(LOH, accessorial_charge_description) %>% 
  filter(accessorial_charge_description %in% c("FUEL CHARGE",
                                               # "ENERGY CHARGE", "ENERGY SURCHARGE", "ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)", "FUEL", "FUEL OTHER",
                                               "FUEL SURCHARGE"))

stats_df_mode <- scaled_data %>%
  group_by(mode, accessorial_charge_description) %>%
  summarize(lane_count = n(),
            med_min   = round(median(min), 2),
            med_max   = round(median(max), 2),
            med_avg   = round(median(avg), 2),
            med_med   = round(median(med), 2),
            med_pct_of_base = round(median(pct_of_base), 2),
            med_pct_of_total = round(median(pct_of_total), 2),
            med_range = round(median(range), 2),
            med_var   = round(median(var), 2)
           ) %>% 
  arrange(mode, accessorial_charge_description) %>% 
  filter(accessorial_charge_description %in% c("FUEL CHARGE",
                                               # "ENERGY CHARGE", "ENERGY SURCHARGE", "ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)", "FUEL", "FUEL OTHER",
                                               "FUEL SURCHARGE"))
stats_df_mode_LOH <- scaled_data %>%
  group_by(mode, LOH, accessorial_charge_description) %>%
  summarize(lane_count = n(),
            med_min   = round(median(min), 2),
            med_max   = round(median(max), 2),
            med_avg   = round(median(avg), 2),
            med_med   = round(median(med), 2),
            med_pct_of_base = round(median(pct_of_base), 2),
            med_pct_of_total = round(median(pct_of_total), 2),
            med_range = round(median(range), 2),
            med_var   = round(median(var), 2)
           ) %>% 
  arrange(mode, accessorial_charge_description) %>% 
  filter(accessorial_charge_description %in% c("FUEL CHARGE",
                                               # "ENERGY CHARGE", "ENERGY SURCHARGE", "ENERGY SURCHARGE (FUEL ADJUSTMENT FACTOR)", "FUEL", "FUEL OTHER",
                                               "FUEL SURCHARGE"))

```

# Elbow plot

```{r}
elbowplot <- function(data, nc=15, seed=123) {
  elbow <- (nrow(data)-1) * sum(apply(data,2,var))
  for (i in 2:nc) {
    set.seed(seed)
    elbow[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, elbow, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")
}

elbowplot(scaled_data[6:13])
```

# Visualize Kmeans clustering

```{r}
km = kmeans(scaled_data[6:13], 4)
autoplot(km, scaled_data, frame=TRUE)
```

# Identify which cluster each group is in

```{r}
find_mode <- function(x) {
  u   <- unique(x)
  tab <- tabulate(match(x, u))
  u[tab == max(tab)]
}

clustered_data <- scaled_data %>% 
  cbind(., closest_cluster=km$cluster) %>% 
  group_by(accessorial_charge_description) %>%
  mutate(cluster = find_mode(closest_cluster)) %>% 
  distinct(accessorial_charge_description, cluster)

print(clustered_data)

```

# Regression

```{r}
lm_fuelcharge <- use_data %>% 
  filter(accessorial_charge_description == "FUEL CHARGE") %>% 
  lm(formula = accessorial_charge_amount ~ factor(LOH), data = .)
summary(lm_fuelcharge)
```
```{r}
lm_fuelsurcharge <- use_data %>% 
  filter(accessorial_charge_description == "FUEL SURCHARGE") %>% 
  lm(formula = accessorial_charge_amount ~ factor(LOH), data = .)
summary(lm_fuelsurcharge)
```
```{r}
lm_stats_fuelsurcharge <- stats_data %>% 
  filter(accessorial_charge_description == "FUEL SURCHARGE") %>% 
  lm(formula = med ~ factor(LOH), data = .)
summary(lm_stats_fuelsurcharge)
```

